---
title: "Lower Salmon Clusters, Communities, and Home Ranges"
date: 
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mapview)
library(dplyr)
library(dbplyr)
library(sf)
library(sp)
library(RColorBrewer)
library(RSQLite)
library(knitr)
library(flexdashboard)
library(stringr)
library(dendextend)
```

```{r inputs, echo=FALSE}
###### user supplied inputs ###################################################
# path to db
  #dbpath <- "/Users/scottp/DocumentsNew/BighornSheep/FY22-WSF-GIA/Databases/BHS_TriState.db"
  dbpath <- "/Users/scottp/Library/CloudStorage/Box-Box/WSF-GIA-FY22/Database/BHS_TriState.db"

# gps, serology, bacteriology table names in DB
  gps.tab.name <- "AnimalID_GPS_Data_AllCollars_2023_05_10"
  pcr.tab.name <- "TriState_Bacteriology"
  ser.tab.name <- "TriState_PCR_Status"

# Forcing analysis to one herd #
  herd <- "Lower Salmon"  # choose from "Lookout Mountain", "Burnt River", "Yakima Canyon", "Cleman Mountain",
                           #   "Lower Salmon", "Lower Panther Main Salmon"
  
# do we want to subset the data to a biological year or date range?
  bioyear.sub <- TRUE
  bioyear <- 2022

# do we want to subset on a range of dates
  daterange.sub <- FALSE
  daterange.sub.dates <- as.POSIXct(c("2022-01-01 00:01","2022-04-01 00:01"),tz="UTC")
  drtext <- as.character(strptime(daterange.sub.dates, format="%Y-%m-%d",tz=''))
  
# Home range settings
  HRestimator <- "BB"   # must be either "Normal" for bivariate normal or "BB" for brownian bridge
                        #   note that BB will add significant computational time
  contour.percent <- 75   # the input to getverticesHR function to cut the UD
  min.fixes <- 60 # one month of 13-hr data = 60, must be >5
  
# Define projected coord systems for each herd or area #
  # we need a mix as some older packages use the CRS() function which sf does not
  or.prj <- 26911 # UTM11N NAD83
  or.crsprj <- CRS("+init=epsg:26911")  
  wa.crsprj <- CRS("+init=epsg:2285")  # WA state plane north is 2285, south is 2927
  wa.prj <- 2285
  wgs.crsproj <- CRS("+init=epsg:4326") # CRS used for GPS data files
  wgs.proj <- 4326
  id.prj <- 8826 # Idaho Transverse Mercator
  id.crsprj <- CRS("+init=epsg:8826") 
  
```

```{r functions, echo=FALSE}
# Helper file full of longer functions
source("HomeRange-helpers.R")

# local functions to help #
removeMissingGPS <- function(gps){
  # check for missing lat/lon and drop #
  missing.loc <- which(is.na(gps$Latitude),arr.ind=TRUE)
  if (length(missing.loc) > 0) gps <- gps[-(missing.loc),]
  return(gps)
}
```

# Intro {.sidebar}

This dashboard shows some of the workflow and R code behind it, 
although most of the code is hidden in the backdrop. The complete
Markdown .Rmd is available on Github.

Each tab has more in-depth coverage of:

* Clustering by animal GPS locations
* Computing individual home ranges and overlap
* Deriving network clusters from overlap of home ranges
* Linking *Movi* testing data back to the network

Email any questions to:
scott.peckham78@gmail.com


# Simple Clustering
### Evaluation Parameters
```{r printparam, echo=FALSE}
# Print to user what we are doing #
  paste("Evaluating clusters for",herd)
  paste("Subset by Biological Year:",bioyear.sub)
    if(bioyear.sub) paste("Biological Year:",bioyear)
  paste("Subset by date range:",daterange.sub)
   if(daterange.sub) paste("Start Date:",daterange.sub.dates[1], "  End Date:",daterange.sub.dates[2] )
  paste("From table:", gps.tab.name)
```

### GPS Clusters
First step is to evaluate any social organization of collared animals by looking at their GPS locations. Some herds have more obvious population substructure than others. This is a pretty basic analysis but might be helpful for some high-level grouping. However, it may not work really well for animals that show multi-modal type distribution patterns in their location data (i.e. multiple distinct activity centers). 

To do this, we'll read data from the data base, subset to our specified parameters (bio year, date range) and then compute the median location for each animal over the time period of interest. 


```{r body, echo=FALSE}
# Connect to the data base read GPS table we need #
  con <- dbConnect(RSQLite::SQLite(),dbpath, extended_types=TRUE)

    # Direct Read table of testing data
    pcr <- dbReadTable(con,pcr.tab.name)
    ser <- dbReadTable(con,ser.tab.name)
    
    # query for gps, don't read into memory
      gps_db <- tbl(con, gps.tab.name) # reference to the table
      # store in data frame
      gps <- gps_db %>% filter(Herd==herd) %>% collect()
    
      
# close out DB connection
  dbDisconnect(con)

  

```

```{r prep, echo=FALSE}

# make necessary subsets #
  if (bioyear.sub) {
    # add a bioyear column if needed #
    gps <- addBioYear(gps)
    gps <- gps %>% filter(BioYear==bioyear) 
  }
  
  if (daterange.sub) gps <- gps %>% filter(acquisitiontime >= daterange.sub.dates[1] &
                                                  acquisitiontime <= daterange.sub.dates[2])
# check for missing lat/lon and drop #
  gps <- removeMissingGPS(gps)

```

```{r project, echo=FALSE}
# assign spatial ref, subset
  gps.sf <- st_as_sf(gps, coords = c("Longitude", "Latitude"), crs = 4326)

# transorm projection, convert to spatial pts df
    # determine appropriate projected coord system for function input#
  projection <- switch(herd,
            "Burnt River"=or.prj, "Lookout Mountain" =or.prj,
            "Yakima Canyon" = wa.prj, "Cleman Mountain" = wa.prj,
            "Lower Salmon" = id.prj, "Lower Panther Main Salmon" = id.prj)
  crs.projection <- switch(herd,
            "Burnt River"=or.crsprj, "Lookout Mountain" =or.crsprj,
            "Yakima Canyon" = wa.crsprj, "Cleman Mountain" = wa.crsprj,
            "Lower Salmon" = id.crsprj, "Lower Panther Main Salmon" = id.crsprj)
  
  gps.pr <- gps.sf %>% subset(select=AnimalID) %>% st_transform(projection)
  gps.sp <- as(gps.pr,"Spatial")

```

### Viewing our GPS locations

```{r mappoints, echo=FALSE}
nanimal <- length(unique(gps.sf$AnimalID))
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
    col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
    mycolors1 <- sample(col_vector, nanimal)
    
mapview(gps.sf, zcol="AnimalID", legend=TRUE, cex=2.5,lwd=1, col.regions=mycolors1,alpha.regions=0.9)
```

```{r dataprep, echo=FALSE}
# construct the matrix for the function dist()
  df <- cbind(gps.sp@data,gps.sp@coords)
  df.m <- df %>% group_by(AnimalID) %>% summarise(medX=median(coords.x1),
                                                 medY=median(coords.x2),
                                                 sdX=sd(coords.x1),
                                                 sdY=sd(coords.x2))
  df.m <- df.m %>% data.frame()
  xy <- df.m[,2:3]
  rownames(xy) <- df.m$AnimalID
  
```

### Computing Clusters

Next we'll then compute a distance matrix based off median locations and apply a hierarchical clustering method using 'hclust' and the function 'cutree' to determine the number of clusters based off the height input to 'cutree'. The 'height' parameter cuts the cluster dendrogram at a specific value rather than specifying a set number of clusters through the 'k' parameter. In this analysis, we could modify our value of 'd' below based on prior knowledge or what distance we want to consider a minimum for cluster membership. Here 'd' is set to 3x the median standard deviation in the location data when grouped by individual animal. Conversely we could use several tests to optimize the choice of k for each set of data and clustering agglomeration method. In exploratory analyses, the UPGMA or "average" method provided the best fit.

```{r cluster}
# perform clustering
  p.dist <- dist(xy)
  chc <- hclust(p.dist, method="average")
  
  xy.sp <- SpatialPointsDataFrame(matrix(c(xy$medX,xy$medY), ncol=2), 
                                  data.frame(AnimalID=rownames(xy)), proj4string=crs.projection)
  
  
# Distance threshold, larger value will yield fewer clusters
#   6-7k chosen here, it's ~ axis of typical Lookout Mountain home range
  #d <- 6000
  d <- 3 * median(sqrt(df.m$sdX^2+df.m$sdY^2))
  chc.d <- cutree(chc, h=d)
  nclust <- max(chc.d) 
  
# Join results to display sp points
  xy.sp@data <- data.frame(xy.sp@data, Clust=chc.d)
 
# Cluster membership, ordered
  rownames(xy.sp@data) <- NULL
  clusters <- sort(unique(xy.sp@data$Clust))
  members <- c(rep("",length(clusters)))
  for (i in 1:length(clusters)){
    membs <- xy.sp@data$AnimalID[xy.sp@data$Clust==clusters[i]]
    members[i] <- str_flatten(membs,collaps=", ")
  }
  kable(data.frame(Cluster=clusters,Members=members),align='ll')
```

### Plotting Raw Cluster Dendrograms
```{r clusterp1, echo=FALSE,fig.height = 6, fig.width = 8, fig.align = "center"}
#plot(chc, main="Cluster Dendrogram",cex=0.6)

dend <- as.dendrogram(chc)
dend <- set(dend, "labels_cex", 0.6)
plot(dend, main="Cluster Dendrogram")
```

### Plotting Cut Cluster Dendrograms
```{r clusterp2, echo=FALSE,fig.height = 6, fig.width = 8, fig.align = "center"}
nclust <- as.integer(max(xy.sp@data$Clust))
mycolors1 <- sample(col_vector, nclust)
# plot(chc,cex=0.6,main=paste("Cluster Dendrogram cut at ",d," (m)",sep=''))
# rect.hclust(chc,h=d,border=rev(mycolors1))

# color labels by cluster:

col_clus <- mycolors1[chc.d]
labels_colors(dend) <- col_clus[order.dendrogram(dend)]

plot(dend, main=paste("Cluster Dendrogram cut at ",round(d)," (m)",sep=''))
```

### Viewing the Clusters on a Map

Evaluate our results visually:

```{r map, echo=FALSE,fig.height = 6, fig.width = 8, fig.align = "center"}
  
  
  mapview(xy.sp,zcol="Clust",col.regions=mycolors1, alpha.regions=1,label="AnimalID") 

```

# Home Range Overlap
### Computing home ranges

Here we will compute home ranges for every GPS-collared animal in the herd using the 'adehabitatHR' package and either a bivariate normal or brownian bridge kernel function. Then, the amount of overlap between each animal (area or UD) is calculated and stored in a matrix. We've set the kernel function, minimum fixes, and contour level used to compute the home range from the utilization distribution in the user-input section in the head of this Markdown .Rmd

```{r homeranges}
if (HRestimator=="BB"){
homeranges <- calculateBBHomerange(gps.sf,min.fixes=min.fixes,contour.percent=contour.percent, output.proj=projection)
} else homeranges <- calculateHomerange(gps.sf,min.fixes=min.fixes,contour.percent=contour.percent, output.proj=projection)
```

### Map of home ranges


```{r maphr, echo=FALSE,fig.height = 7, fig.width = 9, fig.align = "center"}
    qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
    col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
    mycolors2 <- sample(col_vector, length(homeranges$id))
    
    homeranges.sf <- st_as_sf(homeranges)   # for some reason the BB homeranges CRS are screwed up by adehabitat
    st_crs(homeranges.sf) <- projection
    
    makeHomerangeMap(homeranges.sf,mycolors2)
    # mapview(gps.sf, zcol="AnimalID", legend=TRUE, cex=2,lwd=1, col.regions=mycolors2,alpha.regions=0.9)+
    #   mapview(homeranges.sf, zcol="id", burst=TRUE,legend=TRUE, cex=4,lwd=1, col.regions=mycolors2,alpha=0.6)
    #   
    
    

```

### Evaluating amount of overlap between animals
A plot showing the amount of overlap between each pair of animals. This gets pretty messy with large numbers of individuals, so it probably makes more sense to explore the relationships between animals using this measure in a clustering algorithm.

### Home range overlap
```{r ovplot, echo=FALSE,fig.height = 10, fig.width = 10, fig.align = "center"}

  overlap <- calculateHomerangeOverlap(homeranges)
  overlapImagePlot(overlap)

```


# Clustering from Overlap
### Using overlap as a measure of connectivity between animals
In the last tab, we computed a matrix that contained the fraction of each animals home range (by row in the matrix) contained in every other animal in the herd (by columns). Now, treating this as a weighted adjacency matrix we can use tools from the 'igraph' network analysis package to map clusters viewing this data as a directed social network, with the connection between animals weighted by the amount of home range overlap. Note that in a directed network,the connection A to B can be different than B to A, which matches our data. Here we are showing the adjacency matrix clustered using a hierarchical walktrap method and displayed in two plots: 1) plot of the network and 2) a dendrogram. Note that the group colors in the network plot match the leaf text color in the dendrogram.

### Network plot
```{r networkplot, echo=FALSE, fig.height = 8, fig.width = 10.5, fig.align = "center"}
overlapNetworkPlot(overlap)
```

### Dendrogram 
```{r clusterPlot,echo=FALSE, fig.height = 8, fig.width = 10.5, fig.align = "center"}

overlapClusterDend(overlap)

```

### Display cluster membership

```{r members, echo=FALSE, fig.height = 4, fig.width = 6, fig.align = "center"}
df <- getClusterData(overlap)
clusters <- sort(unique(df$Cluster))
  members <- c(rep("",length(clusters)))
  for (i in 1:length(clusters)){
    membs <- df$AnimalID[df$Cluster==clusters[i]]
    members[i] <- str_flatten(membs,collapse=", ")
  }
  kable(data.frame(Cluster=clusters,Members=members),align='ll')
```


# Linking Testing Data to Networks
### Combining testing results with the social network
Now that we have mapped out existing social groups based on overlap in individual home ranges, we can add the results of testing for Movi to our display. By linking each animal back to the testing results records in our database. Note these plots are interactive.

### Network plot with ELISA Status
```{r attplot1, echo=FALSE, fig.height = 8, fig.width = 8, fig.align = "center" }

attributeNetworkPlot(overlap, display="Elisa", gps)
  
```

### Network plot with PCR Status
```{r attplot2, echo=FALSE, fig.height = 8, fig.width = 8, fig.align = "center" }

attributeNetworkPlot(overlap, display="PCR", gps)
  
```
