---
title: "Clusters and Home Ranges"
output: html_document
date: "2023-04-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mapview)
library(dplyr)
library(sf)
library(sp)
library(RColorBrewer)
library(RSQLite)
library(knitr)
```

```{r functions, echo=FALSE}
# local functions to help #
getBioYear <- function(indate){
  year <- as.numeric(strftime(indate, format="%Y",tz="UTC"))
  month <- as.numeric(strftime(indate, format = "%m",tz="UTC"))
  if(month > 4) bioyear <- year else bioyear <- year-1
  return(bioyear)
}
addBioYear <- function(gps.data){
  # create a capture year column #
  gps.data$BioYear <-unlist(lapply(gps.data$acquisitiontime, getBioYear))
  return(gps.data)
}

removeMissingGPS <- function(gps){
  # check for missing lat/lon and drop #
  missing.loc <- which(is.na(gps$latitude),arr.ind=TRUE)
  gps <- gps[-(missing.loc),]
  return(gps)
}
```



## Simple Clustering

First step is to evaluate any social organization of collared animals by looking at their GPS locations. Some herds have more obvious population substructure.

To do this, we'll read data from the data base, subset to our specified parameters (bio year, date range) and then compute the mean location for each animal over the time period of interest. 

```{r inputs, echo=FALSE}
###### user supplied inputs ###################################################
# path to db
  dbpath <- "/Users/scottp/DocumentsNew/BighornSheep/FY22-WSF-GIA/Databases/BHS_TriState.db"

# gps table name in DB
  gps.tab.name <- "AnimalID_GPS_Data_AllCollars_2023_04_06"

# minimum number of GPS fixes needed to compute a home range (has to be >5)
min.fixes <- 60 # one month of 13-hr data = 60

# we need to define projected coord systems for each herd or area #
  or.prj <- CRS("+init=epsg:26911")  # UTM11N NAD83
  #or.prj <- CRS("+init=epsg:2992")  # oregon lambert
  wa.prj <- CRS("+init=epsg:2285")  # WA state plane north is 2285, south is 2927
  wgs.proj <- CRS("+init=epsg:4326") # CRS used for GPS data files

# do we want to subset the data to a biological year or date range?
  bioyear.sub <- TRUE
  bioyear <- 2022

# do we want to subset on a range of dates
  daterange.sub <- FALSE
  daterange.sub.dates <- as.POSIXct(c("2022-01-01 00:01","2022-04-01 00:01"),tz="UTC")
  drtext <- as.character(strptime(daterange.sub.dates, format="%Y-%m-%d",tz=''))

# Subset to a herd? #
  herd.sub <- TRUE
  herd <- "Lookout Mountain"

# Print to user what we are doing #
  paste("Evaluating clusters for",herd)
  paste("Subset by Biological Year:",bioyear.sub)
    if(bioyear.sub) paste("Biological Year:",bioyear)
  paste("Subset by date range:",daterange.sub)
   if(daterange.sub) paste("Start Date:",daterange.sub.dates[1], "  End Date:",daterange.sub.dates[2] )
  paste("From table:", gps.tab.name)
  
```


```{r body}
# Connect to the data base read GPS table we need #
  con <- dbConnect(RSQLite::SQLite(),dbpath, extended_types=TRUE)
  
    # Read table of AnimalID attributed GPS data 
    gps.full <- dbReadTable(con,gps.tab.name) 
  
# close out DB connection
  dbDisconnect(con)

```

```{r prep, echo=FALSE}
# copy the full data set
  gps <- gps.full
  
# make necessary subsets #
  if (bioyear.sub) {
    # add a bioyear column if needed #
    gps <- addBioYear(gps.full)
    gps <- gps %>% filter(BioYear==bioyear) 
  }
  if (herd.sub) {
    gps <- gps %>% filter(Herd==herd) 
  }
  
  if (daterange.sub) gps <- gps %>% filter(acquisitiontime >= daterange.sub.dates[1] &
                                                  acquisitiontime <= daterange.sub.dates[2])
# check for missing lat/lon and drop #
  gps <- removeMissingGPS(gps)

```

```{r project}
# assign spatial ref, subset
  gps.sf <- st_as_sf(gps, coords = c("longitude", "latitude"), crs = 4326)

# transorm projection, convert to spatial pts df
    # determine appropriate projected coord system for function input#
  projection <- switch(herd,
            "Burnt River"=or.prj, "Lookout Mountain" =or.prj,
            "Yakima Canyon" = wa.prj, "Cleman Mountain" = wa.prj)
  
  gps.pr <- gps.sf %>% subset(select=AnimalID) %>% st_transform(projection)
  gps.sp <- as(gps.pr,"Spatial")

```

## Viewing our GPS locations

```{r mappoints, echo=FALSE}
nanimal <- length(unique(gps.sf$AnimalID))
mycolors1 <- colorRampPalette(brewer.pal(8, "Dark2"))(nanimal)
mapview(gps.sf, zcol="AnimalID", legend=TRUE, cex=4,lwd=1, col.regions=mycolors1,alpha.regions=0.9)
```

```{r dataprep, echo=FALSE}
# construct the matrix for the function dist()
  df <- cbind(gps.sp@data,gps.sp@coords)
  df.m <- df %>% group_by(AnimalID) %>% summarise(meanX=mean(coords.x1),
                                                 meanY=mean(coords.x2))
  df.m <- df.m %>% data.frame()
  xy <- df.m[,2:3]
  rownames(xy) <- df.m$AnimalID
  
```

## Computing Clusters

Next we'll then compute a distance matrix based off mean locations and apply a hierarchical clustering method using 'hclust' and the function 'cutree' to determine the number of clusters based off the height input to 'cutree'. The 'height' parameter cuts the cluster dendrogram at a specific value rather than specifying a set number of clusters through the 'k' parameter. In this analysis, we could modify our valud of 'd' below based on prior knowledge or what distance we want to consider a minimum for cluster membership.
```{r cluster}
# perform clustering
  p.dist <- dist(xy)
  chc <- hclust(p.dist, method="ward.D2")
  
  xy.sp <- SpatialPointsDataFrame(matrix(c(xy$meanX,xy$meanY), ncol=2), 
                                  data.frame(AnimalID=rownames(xy)), proj4string=projection)

# Distance with a 5000m threshold, larger value will yield fewer clusters
#   5-6k chosen here, it's ~ axis of typical Lookout Mountain home range
  d <- 5000
  chc.d5k <- cutree(chc, h=d) 
  
# Join results to display sp points
  xy.sp@data <- data.frame(xy.sp@data, Clust=chc.d5k)
 
# Cluster membership, ordered
  rownames(xy.sp@data) <- NULL
  xy.sp@data %>% arrange(Clust)
```

## Viewing the Clusters on a Map

Evaluate our results visually:

```{r map, echo=FALSE}
  nclust <- max(xy.sp@data$Clust)
  mycolors <- colorRampPalette(brewer.pal(8, "RdYlGn"))(nclust)
  mapview(xy.sp,zcol="Clust",col.regions=mycolors, alpha.regions=1,label="AnimalID") 

```


